TEMPERATURES AND METALLICITIES FOR 10,000 M DWARFS\\ IN THE \apogee\ SURVEY

### 2019-07-23 paper/draft.tex
explain selections

M dwarfs have enourmous potential for understanding structure and
formation on both Galactic and exoplanetary scales
with their properties and compositions.
However, insufficient physical assumptions behind atmospheric models
have limited abilities at reproducing spectral features in stars of the 
coolest temperature regimes ($T_{\rm eff} < 4200\,$K) and further 
exploiting the information content and statistical extent of
current and upcoming large-scale spectroscopic surveys.
Here we present a catalog of spectroscopic temperatures, metallicities,
spectral types for 10,311 M dwarfs in the \apogee\ survey using
\thecannon: a flexible, data-driven spectral-modeling and
parameter-inference framework demonstrated to estimate
stellar-parameter labels ($\teff$, $\logg$, $\feh$, and detailed
abundances) to high precision.
Using a training sample of 87 M dwarfs with optically derived labels spanning 
$2860 < \teff < 4130$\,K callibrated with bolometric temperatures, 
$-0.5 < \feh < 0.5$\,dex callibrated with FGK wide binary metallicities, 
we train a two-parameter model with predictive accuracy (in cross-validation) 
to 77\,K and 0.09\,dex respectively.
We also train a one-dimensional spectral type model using 51 M dwarfs
with known SDSS optical spectral types ranging M0 to M9, to predictive accuracy of
0.9 types.
\color{gcolor}{From the overlap of \textsl{APOGEE DR14} and \textsl{Gaia DR2} we
compare our \cannon\ derived $\teff$ against color--temperature
relations and \cannon\ derived $\feh$ against isochrone models in
\gaia\ color-magnitude space.
Finally we use our results to
estimate stellar radii from absolute magnitudes and \cannon\ temperatures [AND WE FIND WHAT??].}\color{black}
\end{abstract}

@@

\begin{figure*}
	\includegraphics[width=\linewidth]{figures/cmd_selection.png}
	\caption{\gaia\ and \zmass\ color-magnitude cuts for 12,037 sources with $\chi^2<100,000$. Over plotted with orange triangles are the 67 out of 87 training sample sources which have parallaxes measured by \gaia.}
	\label{fig:cmd_selection}
\end{figure*}

@@

\begin{enumerate}
\item \textbf{Known contaminant cut:} We remove known pre-main sequence sources from \citealt{Cottaar:2014} (280 sources), and known binaries from \citealt{ElBadry:2018} and \citealt{Skinner:2018} (58 sources).

\item \textbf{K dwarf cut:} In \zmass\ photometry we select $M_K > 4.3$ to remove likely K dwarfs, which appear reddened in \gaia\ $BP$, $RP$ photometry.

\item \textbf{Pre-main sequence, multiple system and subdwarf cut:} in \gaia\ and \zmass\ we apply $.7<J-K<1.1$ and points in the selction region within the points $\left\{ (1.4, 7.5), (2.2, 7.5), (4.2, 14), (3.3, 14) \right\}$ of ($BP-RP$, $M_G$) to cut remaining likely pre-main sequence, multiple sources above the main sequence and subdwarfs below the main sequence.

\item \textbf{Quality of fit cut:} We apply a $\chi^2$ cut of less than 100,000.

\item \textbf{Training bound selection:} Because extrapolation with \thecannon\ is unreliable, we select only sources with labels inside the range of the training sample: for the reported temperatures and metallicities we impose $2850<T_{\rm eff}<4150$ and $-0.5<\feh<0.5$, and for reported spectral subtypes we impose $0<SPT<9$.
\end{enumerate}

>>>>

M dwarfs have enourmous potential for understanding structure and
formation on both Galactic and exoplanetary scales
with their properties and compositions.
However, insufficient physical assumptions behind atmospheric models
have limited abilities at reproducing spectral features in stars of the 
coolest temperature regimes ($T_{\rm eff} < 4200\,$K) and fully
exploiting the information content and statistical extent of
current and upcoming large-scale spectroscopic surveys.
Here we present a catalog of spectroscopic temperatures, metallicities,
spectral types for 10,311 M dwarfs in the \apogee\ survey using
\thecannon: a flexible, data-driven spectral-modeling and
parameter-inference framework demonstrated to estimate
stellar-parameter labels ($\teff$, $\logg$, $\feh$, and detailed
abundances) to high precision.
Using a training sample of 87 M dwarfs with optically derived labels spanning 
$2860 < \teff < 4130$\,K callibrated with bolometric temperatures, 
$-0.5 < \feh < 0.5$\,dex callibrated with FGK wide binary metallicities, 
we train a two-parameter model with predictive accuracy (in cross-validation) 
to 77\,K and 0.09\,dex respectively.
We also train a one-dimensional spectral type model using 51 M dwarfs
with known SDSS optical spectral types ranging M0 to M9, to predictive accuracy of
0.9 types.
\color{gcolor}{From the overlap of \textsl{APOGEE DR14} and \textsl{Gaia DR2} we
apply our models and compare our \cannon\ derived $\teff$ against color--temperature
relations and \cannon\ derived $\feh$ against isochrone models in
\gaia\ color-magnitude space.
Finally we use our results to
estimate stellar radii from absolute magnitudes and \cannon\ temperatures [AND WE FIND WHAT??].}\color{black}
\end{abstract}

@@

\begin{figure*}
	\includegraphics[width=\linewidth]{figures/cmd_selection.png}
	\caption{\gaia\ and \zmass\ color-magnitude cuts for the12,037 sources with $\chi^2<100,000$. Over plotted with orange triangles are the 67 out of 87 training sample sources which have parallaxes measured by \gaia. The coordinates for the selected quadrangles 
	$\left\{ (1.4, 7.5), (2.2, 7.5), (4.2, 14), (3.3, 14) \right\}$ corresponding to ($BP-RP$, $M_G$) for the \gaia\ color-magnitudes shown left, and 
	$\left\{ (.7, 4.3), (1.1, 4.3), (.7, 9), (1.1, 9) \right\}$ corresponding to ($J-K$, $M_K$) for the \zmass\ color-magnitudes shown right. }
	\label{fig:cmd_selection}
\end{figure*}

@@

\begin{enumerate}
\item \textbf{Quality of fit cut:} We apply a \cannon\ model $\chi^2$ cut of less than 100,000.

\item \textbf{Color-magnitude cuts:} Using \gaia\ and \zmass\ photometries we apply the color-magnitude selections shown in Figure \ref{fig:cmd_selection} to cut sources above the main sequence (which are likely pre-main sequence, reddened K dwarfs and multiple), and subdwarfs below the main sequence.

\item \textbf{Model extraploation cuts:} Because extrapolation with \thecannon\ is unreliable, we select only sources with labels inside the range of the training sample: for the \cannon\ reported temperatures and metallicities we impose $2850<T_{\rm eff}<4150$ and $-0.5<\feh<0.5$, and for \cannon\ reported spectral subtypes we impose $0<SPT<9$.

\item \textbf{Astrometric cut:} \gaia\ $RUWE<1.2$

\end{enumerate}
